# Getting Started with the Template {#sec-getting-started}

## Overview

This template provides a structured workflow for writing a Cardiff University thesis in Quarto and rendering it to PDF using LuaLaTeX. It is intended to give a clear starting point for thesis writing while keeping the writing process simple and reproducible. The template separates content from formatting so that chapters, references, appendices, and front matter can be managed in an organised way.

The thesis is built as a Quarto book project. This means that the main file defines the overall thesis structure, while each chapter is written in a separate `.qmd` file. Front matter elements such as acknowledgements, abbreviations, symbols, and dedication are also stored separately. The overall formatting of the PDF output is controlled through the custom class file and the LaTeX partial files.

This chapter introduces the structure of the template and explains where the main files are located. Chapter 2 then explains how to work with citations, figures, tables, equations, references, and appendices in practice.

## What This Template Includes

The template is organised around a small number of core folders and files.

### Main files

The main project file is `index.qmd`. This contains the thesis metadata and defines the order of the chapters. The bibliography file is `references.bib`, which stores all references cited in the thesis. The custom formatting is controlled through `MastersDoctoralThesis.cls`, together with the LaTeX partials in the `tex/` folder.

### Main folders

The `Chapters/` folder contains the main thesis chapters. Each chapter should be written in its own `.qmd` file. The `Appendices/` folder contains appendix files. The `Frontmatter/` folder contains optional front matter files such as acknowledgements, abbreviations, symbols, and dedication. The `Figures/` folder can be used for static figures that are inserted directly into the thesis.

## How the Thesis Is Structured

The thesis follows a standard structure consisting of front matter, main chapters, references, and appendices. The front matter is generated before the main body begins and includes the title page, abstract, acknowledgements, table of contents, and optional lists such as abbreviations and symbols. The main body consists of the numbered chapters. The references are placed at the end of the thesis, followed by any appendices.

This structure is controlled through the YAML in `index.qmd`, where the order of chapters and appendices is listed explicitly.

## Editing the Metadata

Most of the thesis information is set in the YAML at the top of `index.qmd`. This is where the title, author name, submission date, degree name, university, bibliography file, and chapter order are defined. This is also where you can decide which optional front matter files are included.

As a result, one of the first steps in using the template is to replace the default metadata with your own details and check that the listed chapter files match the structure you intend to use.

## Writing Chapters

Each thesis chapter should be written in a separate `.qmd` file using normal Quarto markdown headings. For example, a chapter should begin with a top-level heading using `#`, followed by sections using `##` and subsections using `###`. Section numbers should not be typed manually because Quarto generates them automatically when section numbering is enabled.

This structure keeps each chapter manageable and makes the full thesis easier to maintain. It also helps ensure that chapter numbering, cross-references, and table of contents entries update correctly when the thesis is re-rendered.

## Why This Template Uses Quarto

Quarto provides a convenient writing environment that combines plain text authoring with high-quality PDF output. It also supports references, figures, tables, equations, and cross-references in a consistent way. For thesis writing, this is especially useful because it allows the thesis to remain modular while still rendering as a single document.

The template therefore aims to provide a practical balance between flexibility and structure. The main goal is not to hide the underlying mechanics completely, but to reduce the amount of formatting work that has to be done while writing.

As with other reproducible writing workflows, the separation of source files, bibliography, and formatting makes revision easier over time [@Reference1; @Reference2].

## Example Citation

The template assumes that your bibliography file contains keys such as `Reference1`, `Reference2`, and so on. An in-text citation can therefore be written as `@Reference1`, while a parenthetical citation with more than one source can be written as `[@Reference1; @Reference2]`.

For example, reproducible document workflows can make large writing projects more manageable [@Reference1; @Reference2].

## Example Equation

Equations can be inserted using standard display-math syntax and labelled for later reference.

$$
y = \alpha + \beta x + \varepsilon
$$ {#eq-example-linear}

Equation @eq-example-linear shows a simple linear specification that can be cross-referenced in the text.

## Example Figure

The following figure is included only as a simple example of chunk-based figures and figure cross-referencing.

```{r}
#| label: fig-template-workflow
#| fig-cap: Example workflow for a Quarto thesis project.
#| fig-width: 6
plot(
  1:5, c(2, 4, 3, 5, 6),
  type = "b",
  xlab = "Stage",
  ylab = "Progress",
  main = "Example Thesis Workflow"
)
```

Figure @fig-template-workflow shows a simple example figure produced directly from an R code chunk.

## Example Table

Tables can also be generated directly in a chapter and labelled for cross-reference.

```{r}
#| label: tbl-template-files
#| tbl-cap: Example overview of core template files.
library(knitr)

kable(
  data.frame(
    File = c("index.qmd", "MastersDoctoralThesis.cls", "references.bib"),
    Purpose = c(
      "Defines the thesis structure and metadata",
      "Controls PDF formatting",
      "Stores bibliography entries"
    )
  )
)
```

Table @tbl-template-files provides a simple example of a table produced in a code chunk.

## Before You Begin Writing

Before drafting chapters, it is worth checking the overall structure of the thesis and ensuring that the project metadata, bibliography file, and chapter order are correct. It is also useful to compile the template once at an early stage to make sure the local Quarto and LaTeX setup is working properly.

Once that is done, the thesis can be developed chapter by chapter within the project structure.

## Concluding Remarks

This chapter has introduced the overall purpose and structure of the template. The next chapter explains the practical workflow in more detail, including how to add chapters, insert figures and tables, cite references, and work with appendices.
